Predicting instabilities of a tuneable ring laser with an iterative map model

### 2020-11-26 Papers/Journal/Tuneable_Lasers.tex
Finished rewriting, proper draft finished!

\begin{abstract}
	We propose a nonlinear iterative map for tuneable ring lasers. Using the generalized nonlinear Schr\"odinger equation, expressions for the transformations undergone by the laser pulse are derived for each of the five components (gain, loss, dispersion, modulation, and nonlinearity) within the laser cavity. These transformations are then iteratively composed to give the overall effect of one round trip of the cavity. We first examine the linear version of the model which we solve analytically. Then the full nonlinear model is investigated numerically. The nonlinear model is able to exhibit wave breaking as well as modulation instability. We highlight the rich landscape and sharp divide for a particular plane of the parameter space.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}

% vary 100 nm ~\cite{bohun2015, burgoyne2010, yamashita2009}

% optical coherence tomography~\cite{bohun2015, burgoyne2014, yamashita2009}, coherent anti-Stokes Raman spectroscopy~\cite{burgoyne2014}, deep tissue multi-photon microscopy~\cite{chung2017}, and diagnostics of ultrafast processes~\cite{burgoyne2014, silfvast2004}

% optical frequency metrology, material processing, terahertz generation \cite{oktem2010}

Tuneable lasers have the ability to vary the frequency of their output by up to about 100 nanometres~\cite{bohun2015, burgoyne2010, yamashita2009}. This ability to vary the frequency leads to applications such as optical coherence tomography~\cite{bohun2015, burgoyne2014, yamashita2009}, coherent anti-Stokes Raman spectroscopy~\cite{burgoyne2014}, deep tissue multi-photon microscopy~\cite{chung2017}, and diagnostics of ultrafast processes~\cite{burgoyne2014, silfvast2004}. Tuneable lasers are usually constructed in a ring with five main components: output coupler, chirped fibre Bragg grating (CFBG), modulator, Er-doped fibre, and pump laser. A typical ring laser cavity is depicted in Fig.~\ref{fig:cavity}. Due to the high power and short duration of the pulses, the Kerr nonlinearity plays a crucial role in the dynamics within the cavity.

Several effects arise within the cavity due to the interplay of dispersion, modulation, and the nonlinearity~\cite{bohun2015, coen1997, lapre2019, meng2020, shao2019, woodward2018}. The two effects of most interest for this paper are wave breaking, and modulation instability. Wave breaking causes the leading edge of a pulse to be red shifted, and the trailing edge blue shifted (in the anomalous dispersion regime, the effect is reversed) to the point that a shock is developed~\cite{anderson1992, rothenberg1989a, rothenberg1989b, tomlinson1984, tomlinson1985}. The shifts in frequencies cause the pulse to become more rectangular in the frequency domain with a linear chirp over most of the pulse. In optics, wave breaking manifests from self-phase modulation (SPM) which is a direct effect of the Kerr nonlinearity causing the pulse to interfere with itself. New frequencies are then generated which induces the rectangular profile~\cite{agrawal2013, woodward2018}. Modulation instability---which typically arises in the anomalous dispersion regime---is the other effect that will be of interest. However, in the presence of multiple frequencies, modulation instability can emerge in the normal dispersion regime as well through cross-phase modulation and four-wave mixing~\cite{agrawal1987, agrawal2013, haelterman1992}. Modulation instability causes a degrade into short pulses, and can cause a laser to lose coherence~\cite{agrawal1987, coen1997, haelterman1992}. Outside of optics, wave breaking and modulation instability occur in other areas with nonlinear waves and dispersive media such as plasma physics, heat propagation, transmission lines, and fluid dynamics~\cite{coen1997, rothenberg1989b}. Within a ring laser cavity wave breaking and modulation instability can become parasitic leading to an unstable and unsustainable pulse. This gives rise to the need of understanding the rich landscape of the parameter space, and determining design principles to guarantee the ring laser is stable and sustainable~\cite{bohun2015, burgoyneemail, finot2008, lapre2019, woodward2018}.

\begin{figure}[tbp]
	\centering
	\input{Cavity}
	\caption{Typical cavity of a fibre ring laser~\cite{burgoyne2014, chung2017, lapre2019, shao2019}. The laser pulses travel clockwise around each loop.}
	\label{fig:cavity}
\end{figure}

The standard equation for studying nonlinear optics is the nonlinear Schr\"odinger equation (NLSE),
\begin{align}
	\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A.
	\label{eq:smallnlse}
\end{align}
Here, $A = A(T, z) : \mathbb{R}^2 \mapsto \mathbb{C}$ is the complex pulse amplitude, $\beta_2 \in \mathbb{R}$ is the second order dispersion, and $\gamma \in \mathbb{R}$ is the coefficient of nonlinearity. In practice, the NLSE, \eqref{eq:smallnlse}, lacks a few key terms, thus, it is often generalized by adding amplification and loss (and occasionally higher order terms). These additions give the generalized nonlinear Schr\"odinger equation (GNLSE)~\cite{agrawal2013, bohun2015, finot2008, peng2018, shtyrina2017, yarutkina2013},
	\begin{align}
	\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A + \frac{1}{2}g(A) A - \alpha A,
	\label{eq:nlse}
\end{align}
where $g(A)$ is an amplifying term due to the gain, and $\alpha \in \mathbb{R}^+$ is the loss due to scattering and absorption.

The GNLSE, \eqref{eq:nlse}, has many applications in nonlinear optics and fibre optic communications, however, it is regularly generalized further to include a modulation term in the context of lasers. This is referred to as the master equation of mode-locking (see \cite{haus1975, haus1984, haus2000, tamura1996, usechak2005}), or sometimes the Ginzburg--Landau equation. Due to the nonlinear term in \eqref{eq:nlse} and the often non-trivial form of $g(A)$, as well as the possible addition of a modulation term, no analytic solution is known. See Haus~\cite{haus2000} for a comprehensive description and history of the theory of mode-locked lasers. The GNLSE well represents the dynamics in, say, a single optical fibre, it is not entirely representative of the underlying physics within a laser cavity. It is assumed in \eqref{eq:nlse} that each process affects the pulse continuously within the cavity. Therefore, the GNLSE, and its derivatives, can be thought of as an `average' model for lasers---this assumption breaks down if the pulse undergoes large changes during one round trip. Often, this is the case due to the specialized, discrete components.

@@

\section{Iterative Map Model}
\label{sec:model}
To resolve some of the issues with an `average' model, it is common to use separate GNLSEs for each component and region in the laser, iteratively passing the pulse from one PDE to the next \cite{lapre2019, meng2020, oktem2010, woodward2018}. This method allows us to better encapsulate specific dynamics introduced by each component. The main drawback of this method is its extreme computational expense, since a nonlinear PDE must be numerically solved for every component and each of the hundreds of round trips in a given simulation. We can alleviate this computational expense by noticing that each effect is localized to its corresponding component in the cavity: essentially all of the dispersion happens within the CFBG~\cite{agrawal2002}, the pulse is only amplified within the Erbium-doped fibre, \emph{et cetera}. Therefore, we can drastically simplify each of the GNLSEs into a form we can solve analytically. We get the benefit of the fine-grain simulation over an `average' model, allowing us to reflect the specific geometry of the cavity, without much additional computational cost.

@@

We continue by deriving our own iterative map model by solving a reduced GNLSE for each component---except modulation, which we consider to be applied externally by an electro-optic modulator. Within the gain fibre, we assume all other effects are negligible, allowing us to simplify \eqref{eq:nlse} to
\begin{align}
	\pdiff{A}{z} &= \frac{1}{2}g(A)A.
	\label{eq:gainexample}
\end{align}
To proceed, we assume the gain has the form
\begin{align}
	g(A) &= \frac{g_0}{1 + E / \Es},& E &= \int_{-\infty}^\infty |A|^2 \, \df T,
	\label{eq:energy}
\end{align}
where $g_0$ is a small signal gain, $E$ is the energy of the pulse, and $\Es$ is the energy at which the gain begins to saturate~\cite{haus1984, shtyrina2017, silfvast2004}. Thus, \eqref{eq:gainexample} becomes
\begin{align}
	\pdiff{A}{z} &= \frac{g_0 A}{2 \left( 1 + E / \Es \right)}.
\end{align}
By assuming the energy entering the gain fibre is $E_g$ and the energy after travelling through the gain fibre of length $L_g$ is $E_{\text{out}}$, we find the energy after amplification is
\begin{align}
	\label{eq:engain}
	E_{\text{out}} &= \Es W \left( \frac{E_g}{\Es} \exp( E_g / \Es ) \exp( g_0 L_g ) \right),
\end{align}
where $W$ is the Lambert $W$ function. Rewriting \eqref{eq:engain} in terms of the pulse gives
\begin{align}
	G(A) &= \left( \frac{\Es}{E_g} W \left( \frac{E_g}{\Es} \exp( E_g / \Es ) \exp( g_0 L_g ) \right) \right)^{1/2} A
	\label{eq:gainform}
\end{align}
for the amplification of the pulse within the Er-doped fibre.

Repeating this same procedure for the loss, dispersion, and nonlinearity yields
\begin{align}
	L(A) &= (1 - R) \exp( - \alpha L_T )A, \\
	D(A) &= \FTi{\exp( i \omega^2 L_D\beta_2/2 ) \FT{A}}, \\
	F(A) &= A \exp( i \gamma |A|^2 L_f ),
\end{align}
where $R$ is the reflectivity of the output coupler (depending on the layout of the laser cavity the loss may instead take the form $L(A) = R \exp( - \alpha L_T )A$), $L_T$ is the total length of the laser circuit, $L_D$ is the length of the dispersive medium, $L_f$ is the length of fibre between the amplifier and the output coupler, and $\mathcal{F}$ denotes the Fourier transform. Finally, we consider the modulation. Here, we assume the pulse is modulated by an electro-optic modulator. This gives the freedom of choosing the profile of the modulation how ever the operator sees fit. For simplicity the representation is taken as the Gaussian
\begin{align}
	M(A) &= \exp( -T^2 / 2 T_M^2 ) A,
	\label{eq:modform}
\end{align}
where $T_M$ is the characteristic width of the modulation. This assumption is not particularly restrictive. Calcaterra and Boldt~\cite{calcaterra2008a} have shown that a sum of Gaussian functions can approximate any square integrable function. Therefore, our solution extends to any continuous, bounded modulation function. Additionally, it is common to use a saturable absorber in place of a modulator. In this case, \eqref{eq:modform} can freely be replaced by~\cite{lapre2019, meng2020, oktem2010, woodward2018}
\begin{align}
	S(A) &= 1 - \frac{q_0}{1 + |A|^2 / P_\text{sat}},
\end{align}
where $q_0$ is the modulation depth, $|A|^2$ is the instantaneous power of the pulse, and $P_\text{sat}$ is the saturation power.

@@

Returning to the process maps \eqref{eq:gainform}--\eqref{eq:modform}, we find the non-dimensional mappings to be (after dropping the tildes)
\begin{align}
	G(A) &= \left(E_g^{-1} W \left( a E_g \textrm{e}^{E_g}\right) \right)^{1/2} A, & F(A) &= A \exp( i b |A|^2), \nonumber \\
	D(A) &= \FTi{\exp( i s^2 \omega^2 ) \FT{A}}, & L(A) &= h A, \label{eq:effects} \\
	M(A) &= \exp( -T^2 / 2 ) A, \nonumber
\end{align}
with the four dimensionless parameters,
\begin{equation}
	\begin{aligned}
		a &= \exp( g_0 L_g ) \sim 8 \times 10^3, & 
		h &= (1 - R) \exp( - \alpha L_T) \sim 0.04, \\
		b &= \gamma L_f \frac{\Es}{T_M} \sim 1, & s &= \left( \frac{\beta_2 L_D}{2 T_M^2} \right)^{1/2} \sim 0.2,
		\label{eq:ndparam}
	\end{aligned}
\end{equation}
as defined by the values in Table~\ref{tab:values}. Here, $a$ is a measure of the strength of the gain fibre, $h$ is the loss due to absorption and the output coupler, $b$ is the ratio of the frequency shift due to the Kerr nonlinearity to the width of the pulse, and $s$ is the ratio of the effectiveness of the CFBG to the width of the pulse. Furthermore, notice that in this non-dimensional form, the modulation has no associated parameter, and each of the other processes in \eqref{eq:ndparam} have their own independent, non-dimensional parameter.

% Table of parameter values
\begin{table*}[tbp]
	\centering
	\caption{Range of parameter values (taken from~\cite{agrawal2013, burgoyne2014, burgoyneemail, tamura1996, usechak2005}).}
 	\label{tab:values}
 	\begin{tabular}{clll}
		\hline\noalign{\smallskip}
		Symbol & Parameter & Value & Units \\
		\hline\noalign{\smallskip}
		$\alpha$ & Absorption of fibre & $10^{-4}$--$0.3$ & m$^{-1}$ \\
		$\beta_2^f$ & Fibre dispersion & $-50$--$50$ & ps$^2$/km \\
		$\gamma$ & Fibre nonlinearity & $0.001$--$0.01$ & W$^{-1}$ m$^{-1}$ \\
		$\beta_2^g L_D$ & Grating dispersion & $10$--$2000$ & ps$^2$ \\
		$L_T$ & Length of cavity & $10$--$100$ & m \\
		\vspace*{-2mm} $L_f$ & Length of fibre between & $0.15$--$1$ & m \\
		& gain and output coupler & & \\
		$L_g$ & Length of gain fibre & $2$--$3$ & m \\
		$T_M$ & Modulation Time & $15$--$150$ & ps \\
		$R$ & Reflectivity of optical coupler & $0.1$--$0.9$ & -- \\
		$\Es$ & Saturation energy & $10^3$--$10^4$ & pJ \\
		$g_0$ & Small signal gain &  $1$--$10$ & m$^{-1}$ \\
		\noalign{\smallskip}\hline
	\end{tabular}
\end{table*}

@@

\begin{figure}[tbp]
	\centering
	\includegraphics{Figures/Evo}
	\caption{Evolution of the envelope during one round trip of the cavity at equilibrium. The pulse decays due to the output coupler, disperses due to the CFBG, is modulated by the modulator, and finally, is amplified by the gain fibre (the envelope is unaltered by the nonlinearity). Note, the $z$ axis is not to scale.}
	\label{fig:cavityevo}
\end{figure}

Now that we have the algebraic effect of each section of the cavity, we are ready to take the composition of the maps to give the effect of one round trip of the cavity. Unlike the models models described in the beginning of this section, the permutation of our components is important as the maps \eqref{eq:effects} do not commute with each other. Furthermore, the order of the components will be dictated by the geometry of a particular laser, however, we the laser is constructed as in Fig.~\ref{fig:cavity},  with the following justifications. The effect of the nonlinearity will me maximized after the pulse passes through the gain fibre---this is the region where the pulse is most energetic. Thus, we take the nonlinearity component to follow the gain. Similarly, we would want the loss component to directly follow the nonlinearity in an attempt to minimize the effect the nonlinearity has on the pulse. This leaves the dispersion and modulation components. For now, let us assume the CFBG precedes the modulator. Finally, we assume the loss component is our reference point since this corresponds to the output coupler, and thus, the point in the cavity we are able to take our measurements. More precisely, let
\begin{align}
	\mathcal{L}(A) = F(G(M(D(L(A)))))
	\label{eq:order}
\end{align}
denote one complete round trip of the cavity. Since a ring laser is cyclic in nature, we use the result of one round trip as the input pulse for the next round trip---restarting the process. We continually repeat this procedure until the envelope and chirp of the pulse reaches equilibrium, or the pulse degrades due to modulation instability. Note that we are uninterested in the phase change as it is not modified by the maps in a meaningful way, and due to the nature of experimental measurements. An example of the evolution of the envelope during one round trip of the laser cavity can be found in Fig.~\ref{fig:cavityevo}.

@@

Also of interest is the energy of the pulse at equilibrium, which we compute numerically by \eqref{eq:energy}. Recall from \eqref{eq:order}, this energy is immediately before passing through the output coupler, and so the energy of each pulse emitted by the laser is $(1 - h^2) E$, and $h^2 E$ remains in the cavity. We plot the energy in the $s$--$b$ plane in Fig.~\ref{fig:energy}. Unsurprisingly, we find the same sharp boundary as we saw in Fig.~\ref{fig:error}. In the unstable region (upper-left) the energy is relatively small. Since the pulse is not sustainable in this region it is not able to foster prominent modes, thus, the intensity is low across the entire pulse. In the stable region (lower-right) we find that the energy smoothly decays as both $b$ and $s$ increase, with the contours being hyperbolic-esque. As $s$ increases, more energy is lost due to modulation since the tails of the pulse become larger. Similarly, as $b$ increases, high frequencies generated by SPM are reduced by our choice of modulation function. From this, we find that a more energetic laser requires weaker dispersion, however, the nonlinearity must be correspondingly low, otherwise modulation instability will destroy the pulse. Moreover, the maximum energy, which occurs in the limit that both $s$ and $b$ go to zero, is slightly larger than 2.5. With our choice of parameters, \eqref{eq:maxenergy} predicts an energy of 2.55 at the origin, inline with Fig.~\ref{fig:energy}.

>>>>

\begin{abstract}
	Simple mathematical models have been unable to predict the conditions that lead to instabilities in a tuneable ring laser. Here, we propose a nonlinear iterative map model for tuneable ring lasers. Solving a reduced nonlinear Schr\"odinger equation for each component, we obtain an algebraic map for each component in the laser cavity, which we then iterate through to give the effect of one complete round trip. By neglecting the nonlinearity initially, we find and analyze the analytic fixed point solution of a linearly chirped Gaussian. We then numerically solve the full nonlinear model, allowing us to probe the underlying interplay of dispersion, modulation, and nonlinearity, as an initial pulse evolves during many round trips of the cavity. In this nonlinear case, we find the chirp saturates, and envelope of the pulse becomes more rectangular in shape. Finally, for a nominal plane in the parameter space, we uncover a rich, sharp boundary separating the stable region and the unstable region where modulation instability degrades the pulse into an unsustainable state.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
Dispersion-tuned actively mode-locked lasers are an active area of optics research since they have many applications owing to their ultrashort pulses, a short as 50 femtoseconds \cite{chung2017}, and an ability to tune their output frequency rapidly~\cite{bohun2015, burgoyne2010, chung2017, yamashita2009}. The capability to alter the operating frequency makes this kind of laser particularly useful for imaging applications, such as optical coherence tomography~\cite{bohun2015, burgoyne2014, yamashita2009}, coherent anti-Stokes Raman spectroscopy~\cite{burgoyne2014}, and deep tissue multi-photon microscopy~\cite{chung2017}. As well as sensing and measuring of other ultrafast processes~\cite{burgoyne2014, silfvast2004, oktem2010}. Tuneable lasers are usually constructed in a ring with five main components: optical coupler, chirped fibre Bragg grating (CFBG), modulator (or saturable absorber), Er-doped fibre, and pump laser. A typical ring laser cavity is depicted in Fig.~\ref{fig:cavity}. Due to the high power and ultrashort duration of the pulses, the Kerr nonlinearity plays a crucial role in the dynamics within the cavity.

Several effects arise within the cavity due to the interplay of dispersion, modulation, and the nonlinearity~\cite{bohun2015, coen1997, lapre2019, meng2020, oktem2010, shao2019, woodward2018}. The two effects of most interest for this paper are wave breaking, and modulation instability. Wave breaking causes the leading edge of a pulse to be red shifted, and the trailing edge blue shifted (in the anomalous dispersion regime, the effect is reversed) to the point that a shock is developed~\cite{anderson1992, rothenberg1989a, rothenberg1989b, tomlinson1984, tomlinson1985}. The shifts in frequencies cause the pulse to become more rectangular in the frequency domain with a linear chirp over most of the pulse. In optics, wave breaking manifests from self-phase modulation (SPM) which is a direct effect of the Kerr nonlinearity causing the pulse to interfere with itself. New frequencies are then generated which induces the rectangular profile~\cite{agrawal2013, woodward2018}. Modulation instability---which typically arises in the anomalous dispersion regime---is the other effect that will be of interest. However, in the presence of multiple frequencies, modulation instability can emerge in the normal dispersion regime as well through cross-phase modulation and four-wave mixing~\cite{agrawal1987, agrawal2013, haelterman1992}. Modulation instability causes a degrade into short pulses, and can cause a laser to lose coherence~\cite{agrawal1987, coen1997, haelterman1992}. Outside of optics, wave breaking and modulation instability occur in other areas with nonlinear waves and dispersive media such as plasma physics, heat propagation, transmission lines, and fluid dynamics~\cite{coen1997, rothenberg1989b}. Within a ring laser cavity wave breaking and modulation instability can become parasitic leading to an unstable and unsustainable pulse. This gives rise to the need of understanding the rich landscape of the parameter space, and determining design principles to guarantee the ring laser is stable and sustainable~\cite{bohun2015, burgoyneemail, finot2008, lapre2019, woodward2018}.

\begin{figure}[tbp]
	\centering
	\input{Cavity}
	\caption{Typical cavity of a fibre ring laser~\cite{burgoyne2014, chung2017, lapre2019, shao2019}. The pump laser travels clockwise around the left-hand loop (red arrows) energizing the Er-doped fibre. The main laser travels clockwise around the right-hand loop (blue arrows) passing through each component, with most of the pulse exiting the cavity through the optical coupler.}
	\label{fig:cavity}
\end{figure}

The canonical equation in nonlinear optics is the nonlinear Schr\"odinger equation (NLSE),
\begin{align}
	\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A.
	\label{eq:smallnlse}
\end{align}
Here, $A = A(T, z)$ is the complex pulse envelope, $T$ is co-moving time, $z$ is the propagation coordinate, and $\beta_2$ and $\gamma$ are the group velocity dispersion and Kerr nonlinearity, respectively. Typically, \eqref{eq:smallnlse} is generalized to incorporate terms more specific to laser physics. A gain term and loss term are added (and sometimes higher order dispersion), to give the generalized nonlinear Schr\"odinger equation (GNLSE)~\cite{agrawal2013, bohun2015, finot2008, peng2018, shtyrina2017, yarutkina2013},
	\begin{align}
	\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A + \frac{1}{2}g(A) A - \alpha A,
	\label{eq:nlse}
\end{align}
where $g(A)$ is an amplifying term due to the gain fibre, and $\alpha$ is the loss due to scattering and absorption.

The GNLSE, \eqref{eq:nlse}, has many applications in nonlinear optics and fibre optic communications, however, it is regularly generalized further to include a modulation term in the context of lasers. The resulting equation is referred to as the master equation of mode-locking (see \cite{haus1975, haus1984, haus2000, tamura1996, usechak2005}), or sometimes the Ginzburg--Landau equation. Due to the nonlinear term in \eqref{eq:nlse} and the often non-trivial form of $g(A)$, as well as the possible addition of a modulation term, no analytic solution is known. See Haus~\cite{haus2000} for a comprehensive description and history of the theory of mode-locked lasers. The GNLSE well represents the dynamics in, say, a single optical fibre, it is not entirely representative of the underlying physics within a laser cavity. It is assumed in \eqref{eq:nlse} that each process affects the pulse continuously within the cavity. Therefore, the GNLSE, and its derivatives, can be thought of as an `average' model for lasers---this assumption breaks down if the pulse undergoes large changes during one round trip. Often, this is the case due to the specialized, discrete components.

@@

\section{Iterative Map Model}
\label{sec:model}
To resolve some of the issues with an `average' model, it is common to use separate GNLSEs for each component and region in the laser, iteratively passing the pulse from one PDE to the next \cite{lapre2019, meng2020, oktem2010, woodward2018}. This method allows us to better encapsulate specific dynamics introduced by each component. The main drawback of this method is its extreme computational expense, since a nonlinear PDE must be numerically solved for every component and each of the hundreds of round trips in a given simulation. We can alleviate this computational expense by noticing that each effect is localized to its corresponding component in the cavity: essentially all of the dispersion happens within the CFBG~\cite{agrawal2002}, the pulse is only amplified within the Er-doped fibre, \emph{et cetera}. Therefore, we can drastically simplify each of the GNLSEs into a form we can solve analytically. We get the benefit of the fine-grain simulation over an `average' model, allowing us to reflect the specific geometry of the cavity, without much additional computational cost.

@@

We continue by deriving our own iterative map model by solving a reduced GNLSE for each component---except modulation, which we consider to be applied externally by an electro-optic modulator. Within the gain fibre, we assume all other effects are negligible, allowing us to simplify \eqref{eq:nlse} to
\begin{align}
	\pdiff{A}{z} &= \frac{1}{2}g(A)A.
	\label{eq:gainexample}
\end{align}
To proceed, we assume the gain has the form
\begin{align}
	g(A) &= \frac{g_0}{1 + E / \Es},& E &= \int_{-\infty}^\infty |A|^2 \, \df T,
	\label{eq:energy}
\end{align}
where $g_0$ is a small signal gain, $E$ is the energy of the pulse, and $\Es$ is the energy at which the gain begins to saturate~\cite{haus1984, shtyrina2017, silfvast2004}, which depends on the power of the pump laser. Thus, \eqref{eq:gainexample} becomes
\begin{align}
	\pdiff{A}{z} &= \frac{g_0 A}{2 \left( 1 + E / \Es \right)}.
\end{align}
By assuming the energy entering the gain fibre is $E_g$ and the energy after travelling through the gain fibre of length $L_g$ is $E_{\text{out}}$, we find the energy after amplification is
\begin{align}
	\label{eq:engain}
	E_{\text{out}} &= \Es W \left( \frac{E_g}{\Es} \exp( E_g / \Es ) \exp( g_0 L_g ) \right),
\end{align}
where $W$ is the Lambert $W$ function. Rewriting \eqref{eq:engain} in terms of the pulse gives
\begin{align}
	G(A) &= \left( \frac{\Es}{E_g} W \left( \frac{E_g}{\Es} \exp( E_g / \Es ) \exp( g_0 L_g ) \right) \right)^{1/2} A
	\label{eq:gainform}
\end{align}
for the amplification of the pulse within the Er-doped fibre.

Repeating this same procedure for the loss, dispersion, and nonlinearity yields
\begin{align}
	L(A) &= (1 - R) \exp( - \alpha L_T )A, \\
	D(A) &= \FTi{\exp( i \omega^2 L_D\beta_2/2 ) \FT{A}}, \\
	F(A) &= A \exp( i \gamma |A|^2 L_f ),
\end{align}
where $R$ is the reflectivity of the optical coupler (depending on the layout of the laser cavity the loss may instead take the form $L(A) = R \exp( - \alpha L_T )A$), $L_T$ is the total length of the laser circuit, $L_D$ is the length of the dispersive medium, $L_f$ is the length of fibre between the amplifier and the optical coupler, and $\mathcal{F}$ denotes the Fourier transform. Finally, we consider the modulation. Here, we assume the pulse is modulated by an electro-optic modulator. This gives the freedom of choosing the profile of the modulation how ever the operator sees fit. For simplicity the representation is taken as the Gaussian
\begin{align}
	M(A) &= \exp( -T^2 / 2 T_M^2 ) A,
	\label{eq:modform}
\end{align}
where $T_M$ is the characteristic width of the modulation. This assumption is not particularly restrictive. Calcaterra and Boldt~\cite{calcaterra2008a} have shown that a sum of Gaussian functions can approximate any square integrable function. Therefore, our solution extends to any continuous, bounded modulation function. Additionally, it is common to use a saturable absorber in place of a modulator. In this case, \eqref{eq:modform} can freely be replaced by~\cite{lapre2019, meng2020, oktem2010, woodward2018}
\begin{align}
	S(A) &= 1 - \frac{q_0}{1 + |A|^2 / P_\text{sat}},
\end{align}
where $q_0$ is the modulation depth, $|A|^2$ is the instantaneous power of the pulse, and $P_\text{sat}$ is the saturation power.

@@

Returning to the process maps \eqref{eq:gainform}--\eqref{eq:modform}, we find the non-dimensional mappings to be (after dropping the tildes)
\begin{align}
	G(A) &= \left(E_g^{-1} W \left( a E_g \textrm{e}^{E_g}\right) \right)^{1/2} A, & F(A) &= A \exp( i b |A|^2), \nonumber \\
	D(A) &= \FTi{\exp( i s^2 \omega^2 ) \FT{A}}, & L(A) &= h A, \label{eq:effects} \\
	M(A) &= \exp( -T^2 / 2 ) A, \nonumber
\end{align}
with the four dimensionless parameters,
\begin{equation}
	\begin{aligned}
		a &= \exp( g_0 L_g ) \sim 8 \times 10^3, & 
		h &= (1 - R) \exp( - \alpha L_T) \sim 0.04, \\
		b &= \gamma L_f \frac{\Es}{T_M} \sim 1, & s &= \left( \frac{\beta_2 L_D}{2 T_M^2} \right)^{1/2} \sim 0.2,
		\label{eq:ndparam}
	\end{aligned}
\end{equation}
as defined by the values in Table~\ref{tab:values}. Here, $a$ is a measure of the strength of the gain fibre, $h$ is the loss due to absorption and the optical coupler, $b$ is the ratio of the frequency shift due to the Kerr nonlinearity to the width of the pulse, and $s$ is the ratio of the effectiveness of the CFBG to the width of the pulse. Furthermore, notice that in this non-dimensional form, the modulation has no associated parameter, and each of the other processes in \eqref{eq:ndparam} have their own independent, non-dimensional parameter.

% Table of parameter values
\begin{table*}[tbp]
	\centering
	\caption{Range of parameter values (taken from~\cite{agrawal2013, burgoyne2014, burgoyneemail, tamura1996, usechak2005}).}
 	\label{tab:values}
 	\begin{tabular}{clll}
		\hline\noalign{\smallskip}
		Symbol & Parameter & Value & Units \\
		\hline\noalign{\smallskip}
		$T_M$ & Characteristic modulation time & $15$--$150$ & ps \\
		$\beta_2^g L_D$ & Dispersion provided by CFBG & $10$--$2000$ & ps$^2$ \\
		$\gamma$ & Kerr nonlinearity coefficient & $0.001$--$0.01$ & W$^{-1}$ m$^{-1}$ \\
		$L_f$ & Length of fibre between gain and optical coupler & $0.15$--$1$ & m \\
		$L_g$ & Length of the Er-doped gain fibre & $2$--$3$ & m \\
		$\alpha$ & Loss of fibre due to scattering and absorption & $10^{-4}$--$0.3$ & m$^{-1}$ \\
		$R$ & Reflectivity of optical coupler & $0.1$--$0.9$ & -- \\
		$\Es$ & Saturation energy of Er-doped gain fibre & $10^3$--$10^4$ & pJ \\
		$\beta_2^f$ & Second order fibre dispersion & $-50$--$50$ & ps$^2$/km \\
		$g_0$ & Small signal gain &  $1$--$10$ & m$^{-1}$ \\
		$L_T$ & Total length of fibre in the cavity & $10$--$100$ & m \\
		\noalign{\smallskip}\hline
	\end{tabular}
\end{table*}

@@

\begin{figure}[tbp]
	\centering
	\includegraphics{Figures/Evo}
	\caption{Evolution of the envelope during one round trip of the cavity at equilibrium. The pulse decays due to the optical coupler, disperses due to the CFBG, is modulated by the modulator, and finally, is amplified by the gain fibre (the envelope is unaltered by the nonlinearity). Note, the $z$ axis is not to scale.}
	\label{fig:cavityevo}
\end{figure}

Now that we have the algebraic effect of each section of the cavity, we are ready to take the composition of the maps to give the effect of one round trip of the cavity. Unlike the models models described in the beginning of this section, the permutation of our components is important as the maps \eqref{eq:effects} do not commute with each other. Furthermore, the order of the components will be dictated by the geometry of a particular laser, however, we the laser is constructed as in Fig.~\ref{fig:cavity},  with the following justifications. The effect of the nonlinearity will me maximized after the pulse passes through the gain fibre---this is the region where the pulse is most energetic. Thus, we take the nonlinearity component to follow the gain. Similarly, we would want the loss component to directly follow the nonlinearity in an attempt to minimize the effect the nonlinearity has on the pulse. This leaves the dispersion and modulation components. For now, let us assume the CFBG precedes the modulator. Finally, we assume the loss component is our reference point since this corresponds to the optical coupler, and thus, the point in the cavity we are able to take our measurements. More precisely, let
\begin{align}
	\mathcal{L}(A) = F(G(M(D(L(A)))))
	\label{eq:order}
\end{align}
denote one complete round trip of the cavity. Since a ring laser is cyclic in nature, we use the result of one round trip as the input pulse for the next round trip---restarting the process. We continually repeat this procedure until the envelope and chirp of the pulse reaches equilibrium, or the pulse degrades due to modulation instability. Note that we are uninterested in the phase change as it is not modified by the maps in a meaningful way, and due to the nature of experimental measurements. An example of the evolution of the envelope during one round trip of the laser cavity can be found in Fig.~\ref{fig:cavityevo}.

@@

Also of interest is the energy of the pulse at equilibrium, which we compute numerically by \eqref{eq:energy}. Recall from \eqref{eq:order}, this energy is immediately before passing through the optical coupler, and so the energy of each pulse emitted by the laser is $(1 - h^2) E$, and $h^2 E$ remains in the cavity. We plot the energy in the $s$--$b$ plane in Fig.~\ref{fig:energy}. Unsurprisingly, we find the same sharp boundary as we saw in Fig.~\ref{fig:error}. In the unstable region (upper-left) the energy is relatively small. Since the pulse is not sustainable in this region it is not able to foster prominent modes, thus, the intensity is low across the entire pulse. In the stable region (lower-right) we find that the energy smoothly decays as both $b$ and $s$ increase, with the contours being hyperbolic-esque. As $s$ increases, more energy is lost due to modulation since the tails of the pulse become larger. Similarly, as $b$ increases, high frequencies generated by SPM are reduced by our choice of modulation function. From this, we find that a more energetic laser requires weaker dispersion, however, the nonlinearity must be correspondingly low, otherwise modulation instability will destroy the pulse. Moreover, the maximum energy, which occurs in the limit that both $s$ and $b$ go to zero, is slightly larger than 2.5. With our choice of parameters, \eqref{eq:maxenergy} predicts an energy of 2.55 at the origin, inline with Fig.~\ref{fig:energy}.
